A Framework for Substructural Type Systems \thanks{James Wood is supported by an EPSRC Studentship. Robert Atkey is supported by EPSRC grant FIXME.}

### 2022-01-30 tex/esop22/bunchedrules.tex
[ edit ] prepare for submission

\Cref{fig:lr-comb} shows our prototypical system restated with
implicit contexts and the bunched combinators. The inductive family is
now denoted $\vdash A$, only mentioning context extensions, as we do
in the rules \AgdaInductiveConstructor{⊸I},
\AgdaInductiveConstructor{⊕E} and \AgdaInductiveConstructor{!E}. Thus,
in the \AgdaInductiveConstructor{var} rule, the context is completely
supressed. The \AgdaInductiveConstructor{⊸I} rule just has to state that a
new variable with usage annotation $\gr1$ and type $A$ is added to the
context. The \AgdaInductiveConstructor{⊸E} rule uses the separating
conjunction ($*$) to combine the premises, indicating that the
usages of the two premises are added together for the
conclusion. The \AgdaInductiveConstructor{⊕E} rule demonstrates the
sharing conjunction $\dot\times$: the scrutinee term $s$ and the
clause terms $t, u$ are combined by separating conjunction, because
their usages must be combined, but the clause terms are combined by
the sharing conjunction, because they have the same usage context.

Bunched combinators, along with supression of unchanged typing
contexts, leads to a more streamlined presentation of the system
without the clutter of explicit usage context
inequalities. However, the larger advantage for us is that systems are
constructed using these combinators \emph{automatically} admit
renaming, substitution, and other scope-, type-, and usage-safe
traversals. If we were to allow arbitrary modification of the context
in premises, these results would not be possible, since there would be
no guarantee that a substitution (for instance) could be ``pushed'' up
from a conclusion to the premises. As we will see in \cref{sec:env},
our generic notion of environment (e.g., a simultaneous substitution)
is based around linear transformations, and so automatically commutes
with the linear combinations of premises induced by the bunched
connectives. This is the key to our generic results for all of the
systems describable in our framework.

>>>>

\Cref{fig:lr-comb} shows our prototypical system restated with
implicit contexts and the bunched combinators. The inductive family is
now denoted $\vdash A$, only mentioning context extensions, as we do
in the rules \AgdaInductiveConstructor{⊸I},
\AgdaInductiveConstructor{⊕E} and \AgdaInductiveConstructor{!E}. Thus,
in the \AgdaInductiveConstructor{var} rule, the context is completely
suppressed. The \AgdaInductiveConstructor{⊸I} rule just has to state that a
new variable with usage annotation $\gr1$ and type $A$ is added to the
context. The \AgdaInductiveConstructor{⊸E} rule uses the separating
conjunction ($*$) to combine the premises, indicating that the
usages of the two premises are added together for the
conclusion. The \AgdaInductiveConstructor{⊕E} rule demonstrates the
sharing conjunction $\dot\times$: the scrutinee term $s$ and the
clause terms $t, u$ are combined by separating conjunction, because
their usages must be combined, but the clause terms are combined by
the sharing conjunction, because they have the same usage context.

Bunched combinators, along with suppression of unchanged typing
contexts, leads to a more streamlined presentation of the system
without the clutter of explicit usage context
inequalities. However, the larger advantage for us is that systems are
constructed using these combinators \emph{automatically} admit
renaming, substitution, and other scope-, type-, and usage-safe
traversals. If we were to allow arbitrary modification of the context
in premises, these results would not be possible, since there would be
no guarantee that a substitution (for instance) could be ``pushed'' up
from a conclusion to the premises. As we will see in \cref{sec:env},
our generic notion of environment (e.g., a simultaneous substitution)
is based around linear transformations, and so automatically commutes
with the linear combinations of premises induced by the bunched
connectives. This is the key to our generic results for all of the
systems describable in our framework.

### 2022-01-30 tex/esop22/env.tex
[ edit ] prepare for submission

We have now seen how to build data types of intrinsically well typed
and well uasged terms for a given \AgdaRecord{System}. In the next
section, we will define a generic traversal function that assigns a
``semantics'' to each term. Traversals operate on open terms, so they
need a way to assign semantics to variables in a typed and usage
respecting manner. This is the function fulfiled by
\emph{environments}.

>>>>

We have now seen how to build data types of intrinsically well typed
and well usaged terms for a given \AgdaRecord{System}. In the next
section, we will define a generic traversal function that assigns a
``semantics'' to each term. Traversals operate on open terms, so they
need a way to assign semantics to variables in a typed and usage
respecting manner. This is the function fulfilled by
\emph{environments}.

### 2022-01-30 tex/esop22/examples.tex
[ edit ] prepare for submission

A result we will use throughout this section is \emph{reification}.
When we have an index-preserving mapping from usage-checked variables to
$\V$-environments, we can construct environments of the form
$\Gamma \env\V \Gamma$ (identity environments) for all $\Gamma$.
This lets us write the \AgdaFunction{reify} function, which  simplifies our
obligations in giving a \AgdaRecord{Semanitics} by coercing
\AgdaFunction{Kripke} functions into just
\AgdaBound{$\C$}-values in an extended context.

@@

The final piece of sematics we need is a \emph{bang} operator.
We allow the
semantic \emph{bang} to be an arbitrary annotation-indexed functor on
world-indexed relations.
This functor must respect all of the structure on the indices, making it a
graded comonad over multiplication, as well as being lax monoidal at any
particular index \AgdaBound{r}.
These laws are listed in \cref{sec:den-sem-extras}.

>>>>

A result we will use throughout this section is \emph{reification}.
When we have an index-preserving mapping from usage-checked variables to
$\V$-environments, we can construct environments of the form
$\Gamma \env\V \Gamma$ (identity environments) for all $\Gamma$.
This lets us write the \AgdaFunction{reify} function, which  simplifies our
obligations in giving a \AgdaRecord{Semantics} by coercing
\AgdaFunction{Kripke} functions into just
\AgdaBound{$\C$}-values in an extended context.

@@

The final piece of semantics we need is a \emph{bang} operator.
We allow the
semantic \emph{bang} to be an arbitrary annotation-indexed functor on
world-indexed relations.
This functor must respect all of the structure on the indices, making it a
graded comonad over multiplication, as well as being lax monoidal at any
particular index \AgdaBound{r}.
These laws are listed in \cref{sec:den-sem-extras}.

### 2022-01-30 tex/esop22/quant-framework.tex
[ edit ] squash and address todos
[ edit ] prepare for submission

\usepackage[paperheight=235mm, paperwidth=155mm,textwidth=12.2cm,textheight=19.3cm]{geometry}
\usepackage[conor]{agda}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{catchfilebetweentags}
\usepackage{cleveref}
\usepackage{cmll}
\usepackage{ebproof}
\usepackage{mathrsfs}
\usepackage{mathtools}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{newunicodechar}
\usepackage{stmaryrd}
%\usepackage{todonotes}
\usepackage{turnstile}
\usepackage{mathpartir}

@@

\newcommand{\figref}[1]{Figure \ref{#1}}
\newcommand{\secref}[1]{Section \ref{#1}}
\newcommand{\defref}[1]{Definition \ref{#1}}

@@

@@

  \keywords{Formalised syntax \and substructural types \and mechanised metatheory
  \and quantitative typing}
\end{abstract}
%
%
%
% \begin{enumerate}
% \item (Section 1)
%   \begin{enumerate}
%   \item   Motivate intrinsically typed syntax (in Agda)
%     \begin{enumerate}
%     \item Show the ``GADT'' for well scoped syntax
%     \item Show the well typed version
%     \end{enumerate}
%   \item Motivate adding posemiring annotations
%     \begin{enumerate}
%     \item Pointing to the existing work (Brunel et al., Ghica and Smith, Granule, previous of ours, Bernardy and Abel)
%     \item The fact that we can encode lots of other systems (DILL, PD, monotonicity, information flow, differential privacy)
%     \end{enumerate}
%   \end{enumerate}
% \item (Section 2) Background / notation fixing for:
%   \begin{enumerate}
%   \item Partially ordered semirings
%   \item Vectors
%   \item Linear maps and linear relations (as a generalisation of $P \leq Q\Psi$ which is easier to work with)
%   \end{enumerate}
% \item (Section 3) Bunched Typing Rules
%   \begin{enumerate}
%   \item Introduces $\lambda\mathcal{R}$
%   \item Rephrases the typing rules in terms of bunches (as well as implicit contexts)
%   \end{enumerate}
% \item (Section 4) Universe of Generic Usage-annotated syntax
%   \begin{enumerate}
%   \item Premises
%   \item Rules
%   \item Systems
%   \item Example(s)
%   \item Intepretation as Terms
%   \item Examples of Terms; using pattern synonyms (point out that proofs of usage accounting are painful)
%   \end{enumerate}
% \item (Section 5) Example Syntaxes
%   \begin{enumerate}
%   \item $\lambda\mathcal{R}$ (incl DILL, PD, differential privacy, monotonicity)
%   \item Inductive types with a well resourced fold
%   \item Bernardy-Abel style
%   \item First-order (circuitry)
%   \item LNL
%   \item $\lambda\mu\tilde{\mu}$
%   \end{enumerate}
% \item (Section 6) Environments
%   \begin{enumerate}
%   \item Motivate in terms of renaming and substitution for simple types (where it is just a function)
%   \item Motivate the general idea of using linear relations (i.e. transformations + weakening)
%     \begin{enumerate}
%     \item Key idea / slogan : environment = change of basis
%     \end{enumerate}
%   \item Definition of environments
%   \item Examples
%   \item Properties, incl. preservation of linear operations
%   \end{enumerate}
% \item (Section 7) Semantics, or How to apply an Environment to a Term
%   \begin{enumerate}
%   \item Renameability
%   \item $\mathsf{semantics}$
%   \end{enumerate}
% \item (Section 8) Examples of Semantics
%   \begin{enumerate}
%   \item Renaming and Substitution (term-valued)
%   \item Denotational Semantics (relational world-indexed (\texttt{Generic/Linear/Example/WRel.agda}))
%   \item Usage Elaborator
%   \end{enumerate}
% \item (Section 9) Related Work
% \item (Section 10) Conclusion
% \end{enumerate}

@@

{\bf TODO: explain and format properly.}
\AgdaInductiveConstructor{!E$'$} and \AgdaInductiveConstructor{!I$'$} are
variants on \AgdaInductiveConstructor{!E} and \AgdaInductiveConstructor{!I}
with some of the usage contexts given explicitly (as well as the proofs that the
usage contexts are correctly related).
On concrete posemirings, unification is particularly poor at inferring these
usage contexts from the proofs because addition and multiplication are no longer
(judgementally) injective.
\ExecuteMetaData[\HeavyItex]{cojoin-explicit}

>>>>

\usepackage[paperheight=235mm, paperwidth=155mm,textwidth=12.2cm,textheight=19.3cm]{geometry}
\usepackage[conor]{agda}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{catchfilebetweentags}
\usepackage{hyperref}
\usepackage[noabbrev,nameinlink]{cleveref}
\usepackage{cmll}
\usepackage{ebproof}
\usepackage{mathrsfs}
\usepackage{mathtools}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{newunicodechar}
\usepackage{stmaryrd}
%\usepackage{todonotes}
\usepackage{turnstile}
\usepackage{mathpartir}

@@

@@

\newenvironment{proofsketch}{%
  \renewcommand{\proofname}{Proof sketch}\proof}{\endproof}

\def\orcidID#1{\smash{\href{http://orcid.org/#1}{\protect\raisebox{-1.25pt}{\protect\includegraphics{orcid_color.eps}}}}}

@@

  \keywords{Formalised syntax \and substructural types \and mechanised metatheory
  \and quantitative typing}
\end{abstract}
%
%
%
% \begin{enumerate}
% \item (Section 1)
%   \begin{enumerate}
%   \item   Motivate intrinsically typed syntax (in Agda)
%     \begin{enumerate}
%     \item Show the ``GADT'' for well scoped syntax
%     \item Show the well typed version
%     \end{enumerate}
%   \item Motivate adding posemiring annotations
%     \begin{enumerate}
%     \item Pointing to the existing work (Brunel et al., Ghica and Smith, Granule, previous of ours, Bernardy and Abel)
%     \item The fact that we can encode lots of other systems (DILL, PD, monotonicity, information flow, differential privacy)
%     \end{enumerate}
%   \end{enumerate}
% \item (Section 2) Background / notation fixing for:
%   \begin{enumerate}
%   \item Partially ordered semirings
%   \item Vectors
%   \item Linear maps and linear relations (as a generalisation of $P \leq Q\Psi$ which is easier to work with)
%   \end{enumerate}
% \item (Section 3) Bunched Typing Rules
%   \begin{enumerate}
%   \item Introduces $\lambda\mathcal{R}$
%   \item Rephrases the typing rules in terms of bunches (as well as implicit contexts)
%   \end{enumerate}
% \item (Section 4) Universe of Generic Usage-annotated syntax
%   \begin{enumerate}
%   \item Premises
%   \item Rules
%   \item Systems
%   \item Example(s)
%   \item Interpretation as Terms
%   \item Examples of Terms; using pattern synonyms (point out that proofs of usage accounting are painful)
%   \end{enumerate}
% \item (Section 5) Example Syntaxes
%   \begin{enumerate}
%   \item $\lambda\mathcal{R}$ (incl DILL, PD, differential privacy, monotonicity)
%   \item Inductive types with a well resourced fold
%   \item Bernardy-Abel style
%   \item First-order (circuitry)
%   \item LNL
%   \item $\lambda\mu\tilde{\mu}$
%   \end{enumerate}
% \item (Section 6) Environments
%   \begin{enumerate}
%   \item Motivate in terms of renaming and substitution for simple types (where it is just a function)
%   \item Motivate the general idea of using linear relations (i.e. transformations + weakening)
%     \begin{enumerate}
%     \item Key idea / slogan : environment = change of basis
%     \end{enumerate}
%   \item Definition of environments
%   \item Examples
%   \item Properties, incl. preservation of linear operations
%   \end{enumerate}
% \item (Section 7) Semantics, or How to apply an Environment to a Term
%   \begin{enumerate}
%   \item Renameability
%   \item $\mathsf{semantics}$
%   \end{enumerate}
% \item (Section 8) Examples of Semantics
%   \begin{enumerate}
%   \item Renaming and Substitution (term-valued)
%   \item Denotational Semantics (relational world-indexed (\texttt{Generic/Linear/Example/WRel.agda}))
%   \item Usage Elaborator
%   \end{enumerate}
% \item (Section 9) Related Work
% \item (Section 10) Conclusion
% \end{enumerate}

@@

\section{A term written without the usage elaborator}\label{sec:explicit-term}

Below is a fully witten-out version of the example term at the end of
\cref{sec:usage-elaborator}.
\AgdaInductiveConstructor{!E$'$} and \AgdaInductiveConstructor{!I$'$} are
variants on \AgdaInductiveConstructor{!E} and \AgdaInductiveConstructor{!I}
with some of the usage contexts given explicitly (as well as the proofs that the
usage contexts are correctly related).
On concrete posemirings (as in this example), unification is particularly poor
at inferring these usage contexts from the proofs because addition and
multiplication are no longer (judgementally) injective.
\ExecuteMetaData[\HeavyItex]{cojoin-explicit}

\section{Omitted pieces of the denotational semantics}\label{sec:den-sem-extras}

\ExecuteMetaData[\WReltex]{WRel}

\ExecuteMetaData[\WReltex]{IR}
\ExecuteMetaData[\WReltex]{tensorR}
\ExecuteMetaData[\WReltex]{lollyR}

\todo{Define big intersection.}
\ExecuteMetaData[\WReltex]{WRelMor}

\ExecuteMetaData[\WReltex]{Bang}

### 2022-01-30 tex/esop22/semantics.tex
[ edit ] prepare for submission

The definition \AgdaFunction{Kripke}\,$\V$\,$\C$\,$\Delta$ is a kind
of function space that describes a $\C$ value parameterised by
$\Delta$-many additional $\V$s (all correctly typed and usage
annotated). It is used to describe how to go under binders in a
Higher-Order Abstract Syntax style---to go under a binder we must
provide semantic interpretations for all the additional variables:

% When going under binders during a recursion, the context will be extended by some $\Theta$. This means that the current environemnt must be extended with $\Theta$s-worth of $\V$s

@@

We can now state the data required to implement a traversal assigning
semantics to terms. For open families $\V$ and $\C$, interpreting
variables and terms respectively, we assume that $\V$ is renamable,
that $\V$ is embeddable in $\C$, and that we have an algebra for a
layer of syntax, where bound variables are handled using the Kripke
function space:

@@

To implement the new recursor \AgdaFunction{semantics}, we use the standard
recursor, which in one case gives us a variable \AgdaBound{v}, and in the other
gives us a structure of subterms \AgdaBound{M}, each of which is in an extended
context.
To deal with a variable \AgdaBound{v}, we look it
up in the environment \AgdaBound{$\rho$}, then use the
\AgdaField{$\sem{\text{var}}$} field to map the resulting
\AgdaBound{$\V$}-value to a \AgdaBound{$\C$}-value.
To deal with a structure of subtems \AgdaBound{M}, we use the functoriality of
the syntactic structure to consider each subterm separately.
On a subterm, we apply \AgdaFunction{body}, which amounts to a recursive call
to \AgdaFunction{semantics} with an extended environment.
Recall that \AgdaFunction{relocate} (\cref{thm:env-resize}) adjusts the
environment \AgdaBound{$\rho$} to work in the usage contexts of the subterms.

@@

For \AgdaFunction{body}, we are given a subterm \AgdaBound{M}, to
which we want to apply \AgdaFunction{semantics}.  To do so, we need an
extended version of the initial environment \AgdaBound{$\rho$}. We
express this as the generation of a Kripke function that produces the
extended enviroment given interpretations of the fresh variables. We
take \AgdaBound{$\rho$}, which is an environment covering
\AgdaBound{$\Delta$}, and \AgdaBound{$\sigma$}, which is an
environment covering \AgdaBound{$\Theta$}, and glue them together
using the inductive rules for generating environments, after having
renamed \AgdaBound{$\rho$} via \cref{thm:env-ren} to make it fit the
new context \AgdaBound{$\Gamma^+$} (intended to be
\ExecuteMetaData[\Snippetstex]{GT}):

>>>>

The definition \AgdaFunction{Kripke}\,$\V$\,$\C$\,$\Delta$ is a kind
of function space that describes a $\C$ value parametrised by
$\Delta$-many additional $\V$s (all correctly typed and usage
annotated). It is used to describe how to go under binders in a
Higher-Order Abstract Syntax style---to go under a binder we must
provide semantic interpretations for all the additional variables:

% When going under binders during a recursion, the context will be extended by some $\Theta$. This means that the current environment must be extended with $\Theta$s-worth of $\V$s

@@

We can now state the data required to implement a traversal assigning
semantics to terms. For open families $\V$ and $\C$, interpreting
variables and terms respectively, we assume that $\V$ is renameable,
that $\V$ is embeddable in $\C$, and that we have an algebra for a
layer of syntax, where bound variables are handled using the Kripke
function space:

@@

To implement the new recursor \AgdaFunction{semantics}, we use the standard
recursor, which in one case gives us a variable \AgdaBound{v}, and in the other
gives us a structure of subterms \AgdaBound{M}, each of which is in an extended
context.
To deal with a variable \AgdaBound{v}, we look it
up in the environment \AgdaBound{$\rho$}, then use the
\AgdaField{$\sem{\text{var}}$} field to map the resulting
\AgdaBound{$\V$}-value to a \AgdaBound{$\C$}-value.
To deal with a structure of subterms \AgdaBound{M}, we use the functoriality of
the syntactic structure to consider each subterm separately.
On a subterm, we apply \AgdaFunction{body}, which amounts to a recursive call
to \AgdaFunction{semantics} with an extended environment.
Recall that \AgdaFunction{relocate} (\cref{thm:env-resize}) adjusts the
environment \AgdaBound{$\rho$} to work in the usage contexts of the subterms.

@@

For \AgdaFunction{body}, we are given a subterm \AgdaBound{M}, to
which we want to apply \AgdaFunction{semantics}.  To do so, we need an
extended version of the initial environment \AgdaBound{$\rho$}. We
express this as the generation of a Kripke function that produces the
extended environment given interpretations of the fresh variables. We
take \AgdaBound{$\rho$}, which is an environment covering
\AgdaBound{$\Delta$}, and \AgdaBound{$\sigma$}, which is an
environment covering \AgdaBound{$\Theta$}, and glue them together
using the inductive rules for generating environments, after having
renamed \AgdaBound{$\rho$} via \cref{thm:env-ren} to make it fit the
new context \AgdaBound{$\Gamma^+$} (intended to be
\ExecuteMetaData[\Snippetstex]{GT}):
